---
title: "Prediction_cispt"
author: "Caterina Carraro"
date: "110822"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load environment

```{r}
load("/home/caterina/data/analysis/prediction/envs/110822_Prediction_cispt_GH.RData")
```

# Load packages

```{r}
library(dplyr)
library(reshape2)
library(ggplot2)
library(GSVA)
library(ggsci)
library(metaMA)
library(dplyr)
library(glmnet)
library(ROCR)
library(Metrics)
library(RColorBrewer)
library(ggpubr)
```

# Upload HPA datasets and anno

```{r}
# load("/home/caterina/data/analysis/prediction/envs/220622_predstart.RData")
```

# Create list of HPA-annotated cell line IC50s

```{r}
# cell_line_HPA <- readRDS("/home/caterina/data/analysis/prediction/reference_files/cell_line_list.rds")
# 
# cispt_sensitivity_df <- read.csv("/home/caterina/data/analysis/prediction/reference_files/cisptIC50s.csv")
# 
# #subset for cell lines included in the HPA cell line atlas
# cispt_sensitivity_df <- subset(cispt_sensitivity_df, cispt_sensitivity_df$Cell.line %in% cell_line_HPA)
```

# Generate signatures
## Restrict the CvsC DEcall [(logFC>2.5; <-2.5), padj<0.01]

```{r}
# load("/home/caterina/data/analysis/prediction/objects/cispt_DEresults.RData")
# DEresults_CvsC <-DEdf
# 
# #Restrict the CvsC DEcall [(logFC>2.5; <-2.5), padj<0.01]
# 
# restricted <- subset(DEresults_CvsC, DEresults_CvsC$regulation %in% c("up", "down"))
# restricted <- subset(restricted, restricted$log2FoldChange>2.5 | restricted$log2FoldChange<(-2.5))
# restricted <- subset(restricted, restricted$padj<0.01)
# 
# #Get the restricted coding DE up and down in Jurkat
# 
# Controls_resuDEG_BDOWN <- subset(restricted, restricted$regulation %in% c("down"))
# Controls_resuDEG_BDOWN <- subset(Controls_resuDEG_BDOWN, GENETYPE=='protein_coding') 
# Controls_resuDEG_BDOWN <- unique(Controls_resuDEG_BDOWN$SYMBOL)
#  
# Controls_resuDEG_BUP <- subset(restricted, restricted$regulation %in% c("up"))
# Controls_resuDEG_BUP <- subset(Controls_resuDEG_BUP, GENETYPE=='protein_coding') 
# Controls_resuDEG_BUP <- unique(Controls_resuDEG_BUP$SYMBOL)
```

## Prediction signature

```{r}
# #CoCena-based 
# 
# load("/home/caterina/data/analysis/prediction/objects/cocena_genelist_cispt.rda")
# 
# #Intersect CoCena with restricted CvsC DE
# 
# sign_restDE_CoCena_up50 <- intersect(Controls_resuDEG_BUP, q50genes_tokeep)
# sign_restDE_CoCena_down50 <- intersect(Controls_resuDEG_BDOWN, q50genes_tokeep)
# sign_restDE_CoCena_q50 <- list(sign_restDE_CoCena_up50, sign_restDE_CoCena_down50)
# names(sign_restDE_CoCena_q50)[1] <- "up"
# names(sign_restDE_CoCena_q50)[2] <- "down"
```

# Perform GSVA

## Cell lines

```{r}
gsva_cell_line_q50 <- as.data.frame(t(gsva(expr = as.matrix(cell_line_NX), gset.idx.list = sign_restDE_CoCena_q50, method = "gsva")))

gsva_cell_line_q50$fill <- ifelse(gsva_cell_line_q50$up > gsva_cell_line_q50$down, "sens", "not-sens")

gsva_cell_line_q50$sub <- gsva_cell_line_q50$up - gsva_cell_line_q50$down
```

#### Visualize the GSVA

```{r}
labels <- subset(cispt_sensitivity_df, cispt_sensitivity_df$IC50 < 55)
labels <- labels$Cell.line

gsva_cell_line_q50$label <- ifelse(rownames(gsva_cell_line_q50) %in% labels, rownames(gsva_cell_line_q50), " ")
gsva_cell_line_q50$circle <- ifelse(rownames(gsva_cell_line_q50) %in% labels, "yes", "no")

ggplot(gsva_cell_line_q50, aes(x=up, y=down, fill = sub, label=label, color=circle)) +
  geom_point(size = 3, shape=21) + theme_bw() + theme(aspect.ratio = 1) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_fill_viridis_c() + ggtitle("GSVA analysis on cell lines") +       geom_text() + scale_color_manual(values = c("black", "red"))
```
### Evaluate prediction

```{r}
corr_table <- gsva_cell_line_q50[,3:4]
corr_table$Cell.line <- rownames(corr_table)
corr_table$fill <- NULL
corr_table <- merge(corr_table, cispt_sensitivity_df, by = "Cell.line")
corr_table <- corr_table[, c(1,2,6)]
corr_table <- subset(corr_table, corr_table$IC50 < 55)

rownames(corr_table) <- corr_table$Cell.line
corr_table$Cell.line <- NULL

ggscatter(corr_table, x = "sub", y = "IC50",
   color = "black", shape = 21, size = 3, 
   add = "reg.line", label = rownames(corr_table),
   add.params = list(color = "blue", fill = "#999999"), 
   conf.int = TRUE, 
   cor.coef = TRUE,
   cor.coeff.args = list(method = "pearson", label.x = 0.15, label.sep = "\n"), xlim = c(-0.6, 1.3), ylim = c(-20, 50))
```

## Tumors

```{r}
# gsva_tumors_q50 <- as.data.frame(t(gsva(expr = as.matrix(tumors), gset.idx.list = sign_restDE_CoCena_q50, method = "gsva")))

gsva_tumors_q50$fill <- ifelse(gsva_tumors_q50$up > gsva_tumors_q50$down, "sens", "not-sens")

gsva_tumors_q50$sub <- gsva_tumors_q50$up - gsva_tumors_q50$down
```

### Visualize the GSVA

```{r}
ggplot(gsva_tumors_q50, aes(x=up, y=down, color = sub)) +
geom_point(size = 0.5) + 
theme_bw() + theme(aspect.ratio = 1) + 
geom_abline(slope = 1, intercept = -0.4, linetype = "dashed") +
scale_color_viridis_c() + ggtitle("GSVA analysis on tumors")
```

#### Inspect sensitivity in tumor types

```{r}
#Annotate the gsva_tumor_q50
 
anno_gsva_tumors_q50 <- gsva_tumors_q50
anno_gsva_tumors_q50$Sample <- rownames(anno_gsva_tumors_q50)
anno_gsva_tumors_q50 <- merge(anno_gsva_tumors_q50, anno_tumors, by = "Sample")

median(anno_gsva_tumors_q50$sub)
 
plot_order_violin <- anno_gsva_tumors_q50%>%group_by(Cancer)%>%summarise_each(funs(median), sub)
plot_order_violin <- arrange(plot_order_violin, sub)
plot_order_violin_vect <- as.vector(plot_order_violin$Cancer)
anno_gsva_tumors_q50$Cancer <- factor(anno_gsva_tumors_q50$Cancer, levels = plot_order_violin_vect)

ggplot(anno_gsva_tumors_q50, aes(x=Cancer, y=sub, fill=Cancer)) + 
geom_violin() +
stat_summary(fun.y=median, geom="point", size=2, color="red") +
geom_abline(slope = 0, intercept = -0.017, linetype = "dashed") +
scale_fill_viridis_d()
```

# Session Info

```{r}
info <- sessionInfo()

info
```










