---
title: "Prediction_3-CePs.Rmd"
author: "Caterina Carraro"
date: "080822"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Load environment

```{r}
load("/home/caterina/data/080822_Prediction_3-CePs_GH.RData")
```

# Load packages

```{r}
library(dplyr)
library(reshape2)
library(ggplot2)
library(GSVA)
library(ggsci)
library(metaMA)
library(dplyr)
library(glmnet)
library(ROCR)
library(Metrics)
library(RColorBrewer)
library(ggpubr)
```

# Upload HPA datasets and anno

```{r}
# load("/home/caterina/data/220622_predstart.RData")
```

# Generate signatures
## Restrict the CvsC DEcall [(logFC>1; <-1), padj<0.01]

```{r}
# load("/home/caterina/data/3CPs_CvsC_DEresults.RData")
# DEresults_CvsC <- DEdf
# 
# #Restrict the CvsC DEcall [(logFC>1; <-1), padj<0.01]
# 
# restricted <- subset(DEresults_CvsC, DEresults_CvsC$regulation %in% c("up", "down"))
# restricted <- subset(restricted, restricted$log2FoldChange>1 | restricted$log2FoldChange< (-1))
# restricted <- subset(restricted, restricted$padj<0.01)
# 
# #Get the restricted coding DE up and down in BxPC-3
# 
# Controls_resuDEG_BDOWN <- subset(restricted, restricted$regulation %in% c("up"))
# Controls_resuDEG_BDOWN <- subset(Controls_resuDEG_BDOWN, GENETYPE=='protein_coding') 
# Controls_resuDEG_BDOWN <- unique(Controls_resuDEG_BDOWN$SYMBOL)
#  
# Controls_resuDEG_BUP <- subset(restricted, restricted$regulation %in% c("down"))
# Controls_resuDEG_BUP <- subset(Controls_resuDEG_BUP, GENETYPE=='protein_coding') 
# Controls_resuDEG_BUP <- unique(Controls_resuDEG_BUP$SYMBOL)
```

## Prediction signature

```{r}
# #CoCena-based 
# 
# load("/home/caterina/data/hCoCena/working_directory/data/genes_for_sign/tokeep_revised_topK.rda")
# 
# #Intersect CoCena with restricted CvsC DE
# 
# sign_restDE_CoCena_up50 <- intersect(Controls_resuDEG_BUP, q50genes_tokeep)
# sign_restDE_CoCena_down50 <- intersect(Controls_resuDEG_BDOWN, q50genes_tokeep)
# sign_restDE_CoCena_q50 <- list(sign_restDE_CoCena_up50, sign_restDE_CoCena_down50)
# names(sign_restDE_CoCena_q50)[1] <- "up"
# names(sign_restDE_CoCena_q50)[2] <- "down"
```

## Control signature topFC

```{r}
# #Take top logFC only
# 
# Controlstop_resuDEG_BUP <- subset(restricted, restricted$regulation %in% c("down"))
# Controlstop_resuDEG_BUP <- subset(Controlstop_resuDEG_BUP, Controlstop_resuDEG_BUP$GENETYPE == "protein_coding")
# Controlstop_resuDEG_BUP <- arrange(Controlstop_resuDEG_BUP, log2FoldChange)
# #change based on signature size
# Controlstop_resuDEG_BUP <- Controlstop_resuDEG_BUP[1:307, ]
# Controlstop_resuDEG_BUP <- Controlstop_resuDEG_BUP$SYMBOL
# 
# Controlstop_resuDEG_BDOWN <- subset(restricted, restricted$regulation %in% c("up"))
# Controlstop_resuDEG_BDOWN <- subset(Controlstop_resuDEG_BDOWN, Controlstop_resuDEG_BDOWN$GENETYPE == "protein_coding")
# Controlstop_resuDEG_BDOWN <- arrange(Controlstop_resuDEG_BDOWN, desc(log2FoldChange))
# #change based on signature size
# Controlstop_resuDEG_BDOWN <- Controlstop_resuDEG_BDOWN[1:123, ]
# Controlstop_resuDEG_BDOWN <- Controlstop_resuDEG_BDOWN$SYMBOL
# 
# #Create the top DE CvsC signature
# 
# sign_topDE_CvsC <- list(Controlstop_resuDEG_BUP, Controlstop_resuDEG_BDOWN)
# names(sign_topDE_CvsC )[1] <- "up"
# names(sign_topDE_CvsC )[2] <- "down"
```

# Perform GSVA

## Cell lines

```{r}
gsva_cell_line_q50 <- as.data.frame(t(gsva(expr = as.matrix(cell_line_NX), gset.idx.list = sign_restDE_CoCena_q50, method = "gsva")))

gsva_cell_line_q50$fill <- ifelse(gsva_cell_line_q50$up > gsva_cell_line_q50$down, "sens", "not-sens")

gsva_cell_line_q50$sub <- gsva_cell_line_q50$up - gsva_cell_line_q50$down
```

#### Visualize the GSVA

```{r}
labels <- c("HEK 293", "A549", "A-431", "PC-3", "NTERA-2", "SH-SY5Y")
gsva_cell_line_q50$label <- ifelse(rownames(gsva_cell_line_q50) %in% labels, rownames(gsva_cell_line_q50), " ")
gsva_cell_line_q50$circle <- ifelse(rownames(gsva_cell_line_q50) %in% labels, "yes", "no")

ggplot(gsva_cell_line_q50, aes(x=up, y=down, fill = sub, label=label, color=circle)) +
  geom_point(size = 3, shape=21) + theme_bw() + theme(aspect.ratio = 1) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_fill_viridis_c() + ggtitle("GSVA analysis on cell lines - T") +       geom_text() + scale_color_manual(values = c("black", "red"))
```

### Evaluate prediction (Fig. 6 C)

```{r}
#corr_table <- read.csv("./PREDIZIONE.csv")
gsva_cell_line_q50$cell.line <- rownames(gsva_cell_line_q50)
corr_table_q50 <- merge(corr_table, gsva_cell_line_q50, by = "cell.line")
corr_table_q50 <- corr_table_q50[, c(1,2,6)]
rownames(corr_table_q50) <- corr_table_q50$cell.line
corr_table_q50$cell.line <- NULL


ggscatter(corr_table_q50, x = "sub", y = "viab",
   color = "black", shape = 21, size = 3, 
   add = "reg.line", label = rownames(corr_table_q50),
   add.params = list(color = "blue", fill="#999999"), 
   conf.int = TRUE, 
   cor.coef = TRUE, 
   cor.coeff.args = list(method = "pearson", label.x = 0.15, label.sep = "\n"), xlim = c(-0.4, 0.35), ylim = c(10, 100))
```

## Tumors

```{r}
gsva_tumors_q50 <- as.data.frame(t(gsva(expr = as.matrix(tumors), gset.idx.list = sign_restDE_CoCena_q50, method = "gsva")))

gsva_tumors_q50$fill <- ifelse(gsva_tumors_q50$up > gsva_tumors_q50$down, "sens", "not-sens")

gsva_tumors_q50$sub <- gsva_tumors_q50$up - gsva_tumors_q50$down
```

### Visualize the GSVA (Fig. 6 supplement 1 C)

```{r}
ggplot(gsva_tumors_q50, aes(x=up, y=down, color = sub)) +
geom_point(size = 0.5) + 
theme_bw() + theme(aspect.ratio = 1) + 
geom_abline(slope = 1, intercept = -0.4, linetype = "dashed") +
scale_color_viridis_c() + ggtitle("GSVA analysis on tumors")
```

#### Inspect sensitivity in tumor types (Fig. 6 D)

```{r}
#Annotate the gsva_tumor_q50
 
anno_gsva_tumors_q50 <- gsva_tumors_q50
anno_gsva_tumors_q50$Sample <- rownames(anno_gsva_tumors_q50)
anno_gsva_tumors_q50 <- merge(anno_gsva_tumors_q50, anno_tumors, by = "Sample")

median(anno_gsva_tumors_q50$sub)
 
plot_order_violin <- anno_gsva_tumors_q50%>%group_by(Cancer)%>%summarise_each(funs(median), sub)
plot_order_violin <- arrange(plot_order_violin, sub)
plot_order_violin_vect <- as.vector(plot_order_violin$Cancer)
anno_gsva_tumors_q50$Cancer <- factor(anno_gsva_tumors_q50$Cancer, levels = plot_order_violin_vect)

ggplot(anno_gsva_tumors_q50, aes(x=Cancer, y=sub, fill=Cancer)) + 
geom_violin() +
stat_summary(fun.y=median, geom="point", size=2, color="red") +
geom_abline(slope = 0, intercept = -0.004, linetype = "dashed") +
scale_fill_viridis_d()
```

# Randomized control signature

```{r}
#Randomized signature for cell lines 
set.seed(123)
random_genes <- rownames(cell_line_NX)
random_set_up <- sample(random_genes, size = 307, replace=FALSE)
random_set_down <- sample(random_genes, size = 123, replace=FALSE)
random_set <- list(random_set_up, random_set_down)
names(random_set)[1] <- "up"
names(random_set)[2] <- "down"

#Randomized signature for tumors
set.seed(1)
random_genes_tumors <- rownames(tumors)
random_set_tumors_up <- sample(random_genes_tumors, size = 307, replace=FALSE)
random_set_tumors_down <- sample(random_genes_tumors, size = 123, replace=FALSE)
random_set_tumors <- list(random_set_tumors_up, random_set_tumors_down)
names(random_set_tumors)[1] <- "up"
names(random_set_tumors)[2] <- "down"
```

# Perform GSVA

## Cell lines

```{r}
gsva_cell_line_random <- as.data.frame(t(gsva(expr = as.matrix(cell_line_NX), gset.idx.list = random_set, method = "gsva")))

gsva_cell_line_random$fill <- ifelse(gsva_cell_line_random$up > gsva_cell_line_random$down, "sens", "not-sens")

gsva_cell_line_random$sub <- gsva_cell_line_random$up - gsva_cell_line_random$down
```

### Visualize the GSVA

```{r}
gsva_cell_line_random$label <- ifelse(rownames(gsva_cell_line_random) %in% labels, rownames(gsva_cell_line_random), " ")
gsva_cell_line_random$circle <- ifelse(rownames(gsva_cell_line_random) %in% labels, "yes", "no")

ggplot(gsva_cell_line_random, aes(x=up, y=down, fill = sub, label=label, color=circle)) +
  geom_point(size = 3, shape=21) + theme_bw() + theme(aspect.ratio = 1) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_fill_viridis_c() + ggtitle("GSVA analysis on cell lines") +       geom_text() + scale_color_manual(values = c("black", "red"))
```

### Evaluate prediction (Fig. 6 supplement 1 D)

```{r}
gsva_cell_line_random$cell.line <- rownames(gsva_cell_line_random)
corr_table_random <- corr_table
corr_table_random <- merge(corr_table_random, gsva_cell_line_random, by = "cell.line")
corr_table_random <- corr_table_random[, c(1,2,6)]
rownames(corr_table_random) <- corr_table_random$cell.line
corr_table_random$cell.line <- NULL


ggscatter(corr_table_random, x = "sub", y = "viab",
   color = "black", shape = 21, size = 3, 
   add = "reg.line", label = rownames(corr_table_random),
   add.params = list(color = "blue", fill = "lightgray"), 
   conf.int = TRUE, 
   cor.coef = TRUE, 
   cor.coeff.args = list(method = "pearson", label.x = 0.15, label.sep = "\n"), xlim = c(-0.15, 0.2), ylim = c(10, 100))
```

# TopFC control signature

# Perform GSVA

## Cell lines

```{r}
gsva_cell_line_CvsCtop <- as.data.frame(t(gsva(expr = as.matrix(cell_line_NX), gset.idx.list = sign_topDE_CvsC, method = "gsva")))

gsva_cell_line_CvsCtop$fill <- ifelse(gsva_cell_line_CvsCtop$up > gsva_cell_line_CvsCtop$down, "sens", "not-sens")

gsva_cell_line_CvsCtop$sub <- gsva_cell_line_CvsCtop$up - gsva_cell_line_CvsCtop$down
```
### Visualize the GSVA

```{r}
gsva_cell_line_CvsCtop$label <- ifelse(rownames(gsva_cell_line_CvsCtop) %in% labels, rownames(gsva_cell_line_CvsCtop), " ")
gsva_cell_line_CvsCtop$circle <- ifelse(rownames(gsva_cell_line_CvsCtop) %in% labels, "yes", "no")

ggplot(gsva_cell_line_CvsCtop, aes(x=up, y=down, fill = sub, label=label, color=circle)) +
  geom_point(size = 3, shape=21) + theme_bw() + theme(aspect.ratio = 1) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_fill_viridis_c() + ggtitle("GSVA analysis on cell lines") +       geom_text() + scale_color_manual(values = c("black", "red"))
```

### Evaluate prediction (Fig. 6 supplement 1 E)

```{r}
gsva_cell_line_CvsCtop$cell.line <- rownames(gsva_cell_line_CvsCtop)
corr_table_topFC <- merge(corr_table, gsva_cell_line_CvsCtop, by = "cell.line")
corr_table_topFC <- corr_table_topFC[, c(1,2,6)]
rownames(corr_table_topFC) <- corr_table_topFC$cell.line
corr_table_topFC$cell.line <- NULL


ggscatter(corr_table_topFC, x = "sub", y = "viab",
   color = "black", shape = 21, size = 3, 
   add = "reg.line", label = rownames(corr_table_topFC),
   add.params = list(color = "blue", fill = "lightgray"), 
   conf.int = TRUE, 
   cor.coef = TRUE, 
   cor.coeff.args = list(method = "pearson", label.x = 0.15, label.sep = "\n"), xlim = c(-0.7, 0.3), ylim = c(0, 100))
```

# Classifier

## All tumor types
### Define training and test sets

```{r}
anno_ML_tumors <- anno_gsva_tumors_q50
rownames(anno_ML_tumors) <- anno_ML_tumors$Sample

#put anno_ML_tumors rows (Sample) in same order of tumors column for ML input
anno_ML_tumors <- anno_ML_tumors[colnames(tumors),]

set.seed(123)

ts.samples <- sample(rownames(anno_ML_tumors), size = floor(0.8*nrow(anno_ML_tumors))) # 80% in training set
vs.samples <- rownames(anno_ML_tumors)[!rownames(anno_ML_tumors) %in% ts.samples] # rest in validation / testing set
```

### Continuous classification

```{r}
ts <- tumors[,ts.samples]
ts <- t(ts) # samples have to be in columns for glmnet
vs <- tumors[,vs.samples]
vs <- t(vs)
ts.labels <- anno_ML_tumors[ts.samples, "sub"] # training labels
vs.labels <- anno_ML_tumors[vs.samples, "sub"] # validation labels
```

### Training the model

```{r}
set.seed(123)

model.temp <- cv.glmnet(x = ts, y = ts.labels, family = "gaussian", alpha = 1) # cv to find optimal s (penalty parameter lambda)

# this is the actual prediction on the testing data
pred.dist <- predict(model.temp, newx = vs, type = "response", prob = T, s = model.temp$lambda.min) # this gives probabilities for each sample to be
```

### Plot actual vs predicted values (Fig. 6 F)

```{r}
tumor.labels <- anno_ML_tumors[vs.samples, "Cancer"]

result <- data.frame(actual = vs.labels, predicted = pred.dist, Cancer = tumor.labels)
names(result) <-  c("actual", "predicted", "Cancer")

plot_order <- result%>%group_by(Cancer)%>%summarise_each(funs(median), actual)

plot_order <- arrange(plot_order, actual)
plot_order_vect <- as.vector(plot_order$Cancer)
result$Cancer <- factor(result$Cancer, levels = plot_order_vect)

ggplot(result, aes(x = actual, y = predicted, color = Cancer)) + 
  geom_point() +
  theme_bw() + theme(aspect.ratio = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 2) +
  xlim(-0.5, 0.5)+
  ylim(-0.5, 0.5)+
  scale_color_viridis_d()

cor.test(result$actual, result$predicted)
```

### For calculating the error of the model, you can use the root square mean error

```{r}
rmse(result$actual, result$predicted)
```

# Classifier signatureless (SL)

We train and test the classifier excluding genes belonging to the signature

## Assign continuous labels

```{r}
#employ same set of ts and vs samples (and their labels), but remove signature genes from transcriptomes
sign_genes <- c(sign_restDE_CoCena_q50[["up"]], sign_restDE_CoCena_q50[["down"]])
tumors_SL <- subset(tumors, !(rownames(tumors)%in%sign_genes))
ts_SL <- tumors_SL[,ts.samples]
ts_SL <- t(ts_SL) 

vs_SL <- tumors_SL[,vs.samples]
vs_SL <- t(vs_SL)
```

### Training the SL model

```{r}
set.seed(123)

model.temp_SL <- cv.glmnet(x = ts_SL, y = ts.labels, family = "gaussian", alpha = 1) # cv to find optimal s (penalty parameter lambda)

# this is the actual prediction on the testing data
pred.dist_SL <- predict(model.temp_SL, newx = vs_SL, type = "response", prob = T, s = model.temp_SL$lambda.min) # this gives probabilities for each sample to be
```

### Plot actual vs SL predicted values (Fig. 6 supplement 1 F)

```{r}
tumor.labels <- anno_ML_tumors[vs.samples, "Cancer"]

result_SL <- data.frame(actual = vs.labels, predicted = pred.dist_SL, Cancer = tumor.labels)
names(result_SL) <-  c("actual", "predicted", "Cancer")

plot_order_SL <- result_SL%>%group_by(Cancer)%>%summarise_each(funs(median), actual)

plot_order_SL <- arrange(plot_order_SL, actual)
plot_order_SL_vect <- as.vector(plot_order_SL$Cancer)
result_SL$Cancer <- factor(result_SL$Cancer, levels = plot_order_SL_vect)

ggplot(result_SL, aes(x = actual, y = predicted, color = Cancer)) + 
  geom_point() +
  theme_bw() + theme(aspect.ratio = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 2) +
  xlim(-0.5, 0.5)+
  ylim(-0.5, 0.5)+
  scale_color_viridis_d()

cor.test(result_SL$actual, result_SL$predicted)
```

### For calculating the error of the model, you can use the root square mean error

```{r}
rmse(result_SL$actual, result_SL$predicted)
```

# Tumor type-specific classifier
### Define training and test sets

```{r}
cancer_types <- unique(anno_ML_tumors$Cancer)
specific_ML_tumors <- list()

for (i in cancer_types){
   
   dataset <- subset(anno_ML_tumors, anno_ML_tumors$Cancer == i)
   specific_ML_tumors[[paste0(i)]] <- dataset
   
}

TTsets <- list()

set.seed(123)

for (i in cancer_types){

dataset <- specific_ML_tumors[[i]]
ts.samples <- sample(rownames(dataset), size = floor(0.8*nrow(dataset))) # 80% in training set
vs.samples <- rownames(dataset)[!rownames(dataset) %in% ts.samples] # rest in validation / testing set
TTsets[[i]][["ts.samples"]] <- ts.samples
TTsets[[i]][["vs.samples"]] <- vs.samples

}
```

### Continuous classification

```{r}
TTlabels <- list()
TTglmnet_samples <- list()

for (i in cancer_types){
   
ts.samples <- TTsets[[i]][["ts.samples"]]
ts <- tumors[,ts.samples]
ts <- t(ts) # samples have to be in columns for glmnet

vs.samples <- TTsets[[i]][["vs.samples"]]
vs <- tumors[,vs.samples]
vs <- t(vs)
ts.labels <- anno_ML_tumors[ts.samples, "sub"] # training labels
vs.labels <- anno_ML_tumors[vs.samples, "sub"] # validation labels

TTlabels[[i]][["ts.labels"]] <- ts.labels
TTlabels[[i]][["vs.labels"]] <- vs.labels

TTglmnet_samples[[i]][["ts"]] <- ts
TTglmnet_samples[[i]][["vs"]] <- vs

}
```

### Training the model

```{r}
set.seed(123)

tumor_spec_pred <- list()

for (i in cancer_types){

ts <- TTglmnet_samples[[i]][["ts"]]
ts.labels <- TTlabels[[i]][["ts.labels"]]

model.temp <- cv.glmnet(x = ts, y = ts.labels, family = "gaussian", alpha = 1) # cv to find optimal s (penalty parameter lambda)

vs <- TTglmnet_samples[[i]][["vs"]]
# this is the actual prediction on the testing data
pred.dist <- predict(model.temp, newx = vs, type = "response", prob = T, s = model.temp$lambda.min) # this gives probabilities for each sample to be

tumor_spec_pred[[i]] <- pred.dist

}
```

### Plot actual vs predicted values, calculate cor and RMSE (Fig. 6 supplement 2 A)

```{r}
for (i in cancer_types){

vs.labels <- TTlabels[[i]][["vs.labels"]]
pred.dist <- tumor_spec_pred[[i]]
vs.samples <- TTsets[[i]][["vs.samples"]]
tumor.labels <- anno_ML_tumors[vs.samples, "Cancer"]

result <- data.frame(actual = vs.labels, predicted = pred.dist, Cancer = tumor.labels)
names(result) <-  c("actual", "predicted", "Cancer")

plot_order <- result%>%group_by(Cancer)%>%summarise_each(funs(median), actual)

plot_order <- arrange(plot_order, actual)
plot_order_vect <- as.vector(plot_order$Cancer)
result$Cancer <- factor(result$Cancer, levels = plot_order_vect)

prediction_plot <- ggplot(result, aes(x = actual, y = predicted, color = Cancer)) + 
  geom_point() +
  theme_bw() + theme(aspect.ratio = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 2) +
  xlim(-1, 1)+
  ylim(-1, 1)+
  scale_color_viridis_d()

print(i)

print(prediction_plot)

print(cor.test(result$actual, result$predicted))

print(rmse(result$actual, result$predicted))

}
```

# Plot of THCG20 signature genes

### Volcano Plot signature (Fig. 6 B)

```{r}
#Get signature genes
T_sign_up <- sign_restDE_CoCena_up50
T_sign_down <- sign_restDE_CoCena_down50
sign_updown <- c(T_sign_up, T_sign_down)
DF_sign_updown <- as.data.frame(sign_updown)

#Subset for CvsC DE genes
DF_CvsC <- DEresults_CvsC[, c(2,7,9,12,13)]
DF_CvsC <- subset(DF_CvsC, DF_CvsC$regulation %in% c("up", "down"))

#assign label to key genes
labels <- c("HSPA5", "HERPUD1", "YOD1", "SCD", "ERCC6", "AUNIP", "FDFT1", "BNIP3", "TGFBI", "DDIT4", "FABP5", "STUB1", "SLFN11", "CYP3A5", "ADNP2")
DF_CvsC$label <- ifelse(DF_CvsC$SYMBOL %in% labels, DF_CvsC$SYMBOL, " ")
#change logFC sign (BxPC-3 vs HCT-15)
DF_CvsC$log2FoldChange <- DF_CvsC$log2FoldChange*-1

#assign color (grey DE non signature, red or blue DE in signature)
DF_CvsC$color <- ifelse(DF_CvsC$SYMBOL %in% sign_updown, "color", "c_grey")
DF_CvsC$color <- ifelse(DF_CvsC$regulation == "down" & DF_CvsC$color == "color", "b_red", DF_CvsC$color)
DF_CvsC$color <- ifelse(DF_CvsC$regulation == "up" & DF_CvsC$color == "color", "a_blue", DF_CvsC$color)
DF_CvsC <- DF_CvsC %>% group_by(color) %>% arrange(desc(color))
DF_CvsC$color <- factor(DF_CvsC$color, levels=c("a_blue", "b_red", "c_grey"))

ggplot(data=DF_CvsC, aes(x=log2FoldChange, y=-log10(padj), colour=color, label=label)) +
      geom_point(alpha=0.5, size=1.75) +
      scale_color_manual(values=c("cornflowerblue", "red", "grey"))+
      geom_text(colour="black") +
      xlab("log2(FoldChange)") +
      ylab("-log10(padj)") +
      ylim(0, 50) + 
      xlim(-15, 15) + 
      geom_vline(xintercept = 0, colour="black")+
      geom_vline(xintercept = c(1, -1), colour="red")+
      geom_hline(yintercept=-log(0.01, 10),colour="red")+
      theme_bw()+
      theme(aspect.ratio = 1)
```

### Signature enrichment

#### GSEA dotplot for multiple comparisons (Fig. 6 supplement 1 B)

```{r}
dotplotGSEA_MC <- function(x,
                        show=25,
                        font.size=10,
                        title.size=10,
                        title.width=100){

    ggplot(x, aes(x = Cluster, y = Description, color = pvalue)) +
      geom_point(aes(size = Count)) +
      scale_x_discrete(drop = FALSE) +
      scale_colour_gradientn(colours=c('darkorange1',
                                       '#FFFF0000',
                                       '#FFFF0000'),
                             limits=c(0,1),
                             values   = c(0,0.05,0.2,1),
                             breaks   = c(0.05,0.2,1),
                             labels = format(c(0.05,0.2,1))) +
      ylab(NULL) +
      # ggtitle(paste(strwrap(unique(x$Enrichment), width=title.width), collapse = "\n"))+
      theme_bw() +
      theme(text = element_text(size=font.size),
            plot.title = element_text(size=title.size), axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
  }
```

```{r}
# load universe
#load("/home/caterina/data/universe_Entrez.RData")

sign_up_T_entrez <- bitr(sign_restDE_CoCena_up50, fromType = "SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID

sign_up_GO <- enrichGO(gene = sign_up_T_entrez,
          universe = universe_Entrez,
          OrgDb = "org.Hs.eg.db",
          ont = "BP",
          pAdjustMethod = "none",
          pvalueCutoff  = 1,
          qvalueCutoff  = 1,
          readable      = T)

sign_down_T_entrez <- bitr(sign_restDE_CoCena_down50, fromType = "SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")$ENTREZID

sign_down_GO <- enrichGO(gene = sign_down_T_entrez,
          universe = universe_Entrez,
          OrgDb = "org.Hs.eg.db",
          ont = "BP",
          pAdjustMethod = "none",
          pvalueCutoff  = 1,
          qvalueCutoff  = 1,
          readable      = T)

#select top 50 Counts hits
sign_down_GO_cluster <- sign_down_GO@result
sign_down_GO_cluster <- sign_down_GO_cluster[1:50, ]

sign_up_GO_cluster <- sign_up_GO@result
sign_up_GO_cluster <- sign_up_GO_cluster[1:50, ]

#isolate all top terms
down_descr <- sign_down_GO_cluster$Description
up_descr <- sign_up_GO_cluster$Description
descr <- unique(c(down_descr, up_descr))

#filter GO output for top terms
up_GO_filtered <- subset(sign_up_GO, sign_up_GO$Description %in% descr)
up_GO_filtered$Cluster <- "up"
down_GO_filtered <- subset(sign_down_GO, sign_down_GO$Description %in% descr)
down_GO_filtered$Cluster <- "down"

#merge the filtered datasets and arrange by pval
updown_GO_filtered <- rbind(up_GO_filtered, down_GO_filtered)
updown_GO_filtered <- updown_GO_filtered %>% group_by(Cluster) %>% arrange(pvalue)
updown_GO_filtered <- updown_GO_filtered %>% arrange(desc(Cluster))

#skim out interesting GO terms, 10 per group
selected_GO <- updown_GO_filtered[c(3,4,7,13,14,25,27,30,31,40,51,57,100,101,104,109,119,139,142,149,166),]
selected_GO <- selected_GO$Description

#further polish the dataset to be plotted for final GO
updown_GO_filtered <- subset(updown_GO_filtered, updown_GO_filtered$Description %in% selected_GO)

dotplotGSEA_MC(updown_GO_filtered)
```

# Session Info
```{r}
info <- sessionInfo()

info
```
